A novel method for detecting coalescing binaries in near realtime with Advanced LIGO and beyond \editorial{Submit to PRD or CQG?}

### 2011-02-11 appendix.tex
Adding comments and references
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@613 ccdd05d5-56c1-461a-bf98-85f6876523b4

More commonly, the matched filters are implemented using the \textsc{fft} convolution.  This entails applying \textsc{fft}s to blocks of $D$ samples, with $2 n \leq D$, each block overlapping the previous one by $n$ samples.  There are $N/(D-n)$ such blocks.  Modern implementations of the Cooley-Tukey \textsc{fft}, such as the ubiquitous \texttt{fftw}, require about $4 N \lg N$ operations to evaluate a \textsc{dft} of size $N$~\cite{Johnson:2007p9654}.  \editorial{This is more commonly known as ``overlap-save''.  We should find someone else's operation count and cite it.}  A $D$ sample cross-correlation consists of a forward \textsc{fft}, an $D$ sample dot product, and an inverse \textsc{fft}, totaling $8 D \lg N + 2 D$ operations per block.  Per sample, this is $(8 \lg D + 2) / (1 - n/D)$ operations.

The \textsc{fir} filter implementation has the advantage that it has no intrinsic latency, whereas the \textsc{fft} convolution has at least the latency of the \textsc{fft} block size $D \geq 2 n$.  For example, for a $1.4 - 1.4 \, M_\odot$ template with duration $\sim 1 \, \mathrm{ks}$, the \textsc{fft} convolution has a latency $\geq 2 \, \mathrm{ks}$.  However, the \textsc{fir} filter implementation has the disadvantage of much greater overhead per sample than the \textsc{fft} convolution.  For a $1\,\mathrm{ks}$ template sampled at $4096\,\mathrm{Hz}$, the \textsc{fir} implementation requires about about $n / 8 \lg 2 n = 2.2 \times 10^4$ times more operations per sample than the \textsc{fft} implementation.

>>>>

More commonly, the matched filters are implemented using the \textsc{fft} convolution.  This entails applying \textsc{fft}s to blocks of $D$ samples, with $2 n \leq D$, each block overlapping the previous one by $n$ samples.  There are $N/(D-n)$ such blocks.  Modern implementations of the Cooley-Tukey \textsc{fft}, such as the ubiquitous \texttt{fftw}, require about $4 N \lg N$ operations to evaluate a \textsc{dft} of size $N$~\cite{Johnson:2007p9654}.  \editorial{This is more commonly known as ``overlap-save''.  We should find someone else's operation count and cite it.}  A $D$ sample cross-correlation consists of a forward \textsc{fft}, an $D$ sample dot product, and an inverse \textsc{fft}, totaling $8 D \lg N + 2 D$ operations per block.  Per sample, this is $(8 \lg D + 2) / (1 - n/D)$ operations. \editorial{Drew: Why don't we change this to an overlap of $m$ samples so we can see what happens as we increase the overlap to reduce latency.}

The \textsc{fir} filter implementation has the advantage that it has no intrinsic latency, whereas the \textsc{fft} convolution has at least the latency of the \textsc{fft} block size $D \geq 2 n$.  \editorial{Drew: Should the latency be D-n?} For example, for a $1.4 - 1.4 \, M_\odot$ template with duration $\sim 1 \, \mathrm{ks}$, the \textsc{fft} convolution has a latency $\geq 2 \, \mathrm{ks}$.  However, the \textsc{fir} filter implementation has the disadvantage of much greater overhead per sample than the \textsc{fft} convolution.  For a $1\,\mathrm{ks}$ template sampled at $4096\,\mathrm{Hz}$, the \textsc{fir} implementation requires about about $n / 8 \lg 2 n = 2.2 \times 10^4$ times more operations per sample than the \textsc{fft} implementation.

### 2011-02-11 inspiral_svd.tex
Reverting 606
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@607 ccdd05d5-56c1-461a-bf98-85f6876523b4
Adding comments and references
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@613 ccdd05d5-56c1-461a-bf98-85f6876523b4

\author{Kipp Cannon$^1$, Adrian Chapman$^2$, Chad Hanna$^3$, Drew Keppel$^4$, Antony C. Searle$^2$, Alan J. Weinstein$^2$}

\address{$^1$ Canadian Institute for Theoretical Astrophysics, ON, Canada}
\address{$^2$ LIGO Laboratory - California Institute of Technology, Pasadena, CA, USA}
\address{$^3$ Perimeter Institute for Theoretical Physics, Waterloo, ON, Canada}
\address{$^3$ Albert-Einstein-Institut, Max-Planck-Institut f\"{u}r Gravitationphysik, Hannover, Germany}

\begin{abstract}
Conventional matched filter bank methods for the detection of gravitational waves from the inspiral of compact binaries are computationally expensive, have hundreds of seconds of unavoidable intrinsic latency, and require arrays of largely redundant matched filters.  Novel detection methods that are more computationally efficient and have lower latency will be required to realize the full potential of advanced of gravitational wave detectors that are currently under construction.  In this paper, we describe a new detection method that exploits the properties of inspiral waveforms using multi-rate filtering, principal component analysis, and hierarchical detection.  We provide receiver operating characteristics from a prototype search pipeline that is capable of low-latency or near-realtime detection with greatly reduced computational requirements in comparison with previously described methods.
\end{abstract}

>>>>

\author{Kipp Cannon$^1$, Adrian Chapman$^2$, Chad Hanna$^3$, Drew Keppel$^{4,5}$, Antony C. Searle$^2$, Leo Singer$^2$, Alan J. Weinstein$^2$}

\address{$^1$ Canadian Institute for Theoretical Astrophysics, ON, Canada}
\address{$^2$ LIGO Laboratory - California Institute of Technology, Pasadena, CA, USA}
\address{$^3$ Perimeter Institute for Theoretical Physics, Waterloo, ON, Canada}
\address{$^4$ Albert-Einstein-Institut, Max-Planck-Institut f\"{u}r Gravitationphysik, Hannover, Germany}
\address{$^5$ Leibniz Universit\"{a}t Hannover, Hannover, Germany}

\begin{abstract}
Conventional matched filter bank methods for the detection of gravitational waves from the inspiral of compact binaries are computationally expensive, have hundreds of seconds of unavoidable intrinsic latency, and require arrays of largely redundant matched filters.  Novel detection methods that are more computationally efficient and have lower latency will be required to realize the full potential of advanced of gravitational wave detectors that are currently under construction.  In this paper, we describe a new detection method that exploits the properties of inspiral waveforms using multi-rate filtering, principal component analysis, and hierarchical detection.  We provide receiver operating characteristics from a prototype search pipeline that is capable of low-latency or near-realtime detection with greatly reduced computational requirements in comparison with previously described methods. Help me!
\end{abstract}

### 2011-02-11 introduction.tex
Adding comments and references
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@613 ccdd05d5-56c1-461a-bf98-85f6876523b4

\editorial{This is old news, and doesn't seem relevant to me: EM-directed gravitational wave searches don't benefit from low-latency detection pipelines.} Work has already commenced to trigger gravitational-wave searches from 
electromagnetic observations~\cite{triggeredsearches2008}. \textsc{ligo} observations
ruled out the possibility of GRB070201 originating from the merger of a
\textsc{nsns} or \textsc{nsbh} system in the Andromeda galaxy~\cite{GRB070201}.

@@

This question has been investigated and it seems  at least
possible that gravitational-wave detections could provide a region of the
sky to prompt electromagnetic observation followups~\cite{sylvestre2003}.
\editorial{Cite an in-preparation paper of Larry's?} Much work is currently underway in providing the best possible source 
localization from gravitational-wave detector networks~\cite{markowitz2008, raymond2008, cavalier2006}.  Collaborations are currently forming to provide infrastructure for the gravitational-wave
Astronomers to provide targets of opportunity for electromagnetic 
astronomers~\cite{kanner2008}.  Most models that predict coincident
gravitational-wave and electromagnetic observations also predict that the
peak electromagetic fluence will occur soon after 
gravitational-wave emission~\cite{sylvestre2003}.  Thus in order to maximize
the chance of a successful electromagnetic followup the latency of 
gravitational-wave signal analysis must be made minimal.

>>>>

\editorial{This is old news, and doesn't seem relevant to me: EM-directed
gravitational wave searches don't benefit from low-latency detection
pipelines.; Drew: I agree} Work has already commenced to trigger
gravitational-wave searches from electromagnetic
observations~\cite{triggeredsearches2008}. \textsc{ligo} observations ruled out
the possibility of GRB070201 originating from the merger of a \textsc{nsns} or
\textsc{nsbh} system in the Andromeda galaxy~\cite{GRB070201}.

@@

This question has been investigated and it seems  at least possible that
gravitational-wave detections could provide a region of the sky to prompt
electromagnetic observation followups~\cite{sylvestre2003}.  \editorial{Cite an
in-preparation paper of Larry's? Drew: When it is on arXiv, yes.} Much work is
currently underway in providing the best possible source localization from
gravitational-wave detector networks~\cite{markowitz2008, raymond2008,
cavalier2006}.  Collaborations are currently forming to provide infrastructure
for the gravitational-wave Astronomers to provide targets of opportunity for
electromagnetic astronomers~\cite{kanner2008}.  Most models that predict
coincident gravitational-wave and electromagnetic observations also predict
that the peak electromagetic fluence will occur soon after gravitational-wave
emission~\cite{sylvestre2003}.  Thus in order to maximize the chance of a
successful electromagnetic followup the latency of gravitational-wave signal
analysis must be made minimal.

### 2011-02-11 method.tex
Adding comments and references
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@613 ccdd05d5-56c1-461a-bf98-85f6876523b4

We can exploit the time-frequency structure of the templates by processing each time slice at a reduced sample rate.  In the quadrupole approximation~\cite{finn1993}, the frequency of gravitational radiation resulting from the inspiral of two compact objects evolves according to
\begin{equation}
\label{eq:fgw}
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{-t} \right]^{3/8}
\end{equation}
where $\mathcal{M}$ is the chirp mass and $t$ is the time relative to the coalescence of the binary~\cite{kidder1992, blanchet2002, findchirppaper, hanna2009}. \editorial{Way too many citations here.} Typically the template is truncated at some time before coalescence, corresponding to a finite frequency which is often chosen to be the frequency at the innermost stable circular orbit, $f_\mathrm{final} = f_\mathrm{ISCO} = 4400 \, M_\odot / M$.

@@

\editorial{\textsc{mbta} should be cited in the introduction.}  This idea has been demonstrated by the Virgo Collaboration's \textsc{mbta} pipeline~\cite{beauville2006,beauville2008}, which operated in a low-latency mode during \textsc{ligo}'s sixth science run and Virgo's second science run in 2010.

>>>>

We can exploit the time-frequency structure of the templates by processing each time slice at a reduced sample rate.  In the quadrupole approximation~\cite{finn1993}, the frequency of gravitational radiation resulting from the inspiral of two compact objects evolves according to
\begin{equation}
\label{eq:fgw}
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{-t} \right]^{3/8}
\end{equation}
where $\mathcal{M}$ is the chirp mass and $t$ is the time relative to the coalescence of the binary~\cite{findchirppaper, kidder1992, blanchet2002, hanna2009}. \editorial{Way too many citations here.} Typically the template is truncated at some time before coalescence, corresponding to a finite frequency which is often chosen to be the frequency at the innermost stable circular orbit, $f_\mathrm{final} = f_\mathrm{ISCO} = 4400 \, M_\odot / M$.

@@

\editorial{\textsc{mbta} should be cited in the introduction.}  This idea has been demonstrated by the Virgo Collaboration's \textsc{mbta} pipeline~\cite{Marion2004,beauville2006,beauville2008,Buskulic2010}, which operated in a low-latency mode during \textsc{ligo}'s sixth science run and Virgo's second science run in 2010.
